Learning complex stochastic models with invertible neural networks: a likelihood-free Bayesian approach

### 2019-07-16 latex/paper_main.tex
Prepare final notebooks

To assess the performance of our method on the following application examples, we consider a number of different metrics. To assess the precision of point estimates (posterior means), we compute the normalized root mean squared error (NRMSE) and the coefficient of determination ($R^{2}$) between estimated and true parameter values. To assess the recovery of the full posterior, we compute the Kullback-Leibler (KL) divergence \cite{hershey2007approximating} between the true and the approximate distributions for the toy example, and use simulation-based calibration (SBC, \cite{talts2018validating}) for the other examples where the analytic posterior is not available in closed-form. Details for computing the NRMSE, $R^2$, and SCB can be found in the \textbf{SI}. Code for reproducing the results on all following examples, as well as all network implementations, is freely available at: \href{https://github.com/stefanradev93/cINN}{https://github.com/stefanradev93/cINN}.

@@

As a proof-of-concept, we demonstrate the utility of our method in recovering the true analytic posteriors of the regression coefficients of a conjugate Bayesian regression model. To set the stage, assume we have observed a dataset $\boldsymbol{D} = \{(\boldsymbol{x}^{(i)},y^{(i)})\}_{i=1}^{n}$ with $\boldsymbol{x} \in \mathbb{R}^{d}$ and $y \in \mathbb{R}$. We stack all $\boldsymbol{x}^{(i)}$ row-wise in a design matrix $\boldsymbol{X}$ and model the vector of outcomes $\boldsymbol{y}$ as being conditionally Gaussian given $\boldsymbol{X}$, that is, $\boldsymbol{y} \sim \mathcal{N}_{n}(\boldsymbol{X}\boldsymbol{\theta},a^{-1}\boldsymbol{I})$ where $\boldsymbol{\theta} \in \mathbb{R}^{d}$ and $a$ is the precision (inverse noise variance, $a \equiv 1/\sigma_{y}^{2}$). We place a $d$-dimensional diagonal Gaussian prior on the regression coefficients centered at $0$: $\boldsymbol{\theta} \sim \mathcal{N}_{d}(\boldsymbol{0},b^{-1}\boldsymbol{I})$ where $b$ is the precision of the prior ($b \equiv 1/\sigma_{\boldsymbol{\theta}}^{2}$). Thus, the likelihood $p(\boldsymbol{D}|\boldsymbol{\theta})$ admits the following proportionality:
\begin{align*}
p(\boldsymbol{D}|\boldsymbol{\theta}) \propto \exp\left(-\frac{a}{2}\left(\boldsymbol{y}-\boldsymbol{X}\boldsymbol{\theta}\right)^{T}\left(\boldsymbol{y} - \boldsymbol{X}\boldsymbol{\theta}\right)\right) \numberthis 
\end{align*}
Since the prior of $\boldsymbol{\theta}$ is conjugate to the likelihood (both are Gaussian distributions), the posterior of $\boldsymbol{\theta}$ is also Gaussian, given by:
\begin{align*}
p(\boldsymbol{\theta}|\boldsymbol{D}) \propto \exp\left(-\frac{a}{2}\left(\boldsymbol{y}-\boldsymbol{X}\boldsymbol{\theta}\right)^{T}\left(\boldsymbol{y} - \boldsymbol{X}\boldsymbol{\theta}\right)-\frac{b}{2}\boldsymbol{\theta}^{T}\boldsymbol{\theta}\right) \numberthis 
\end{align*}
Therefore, the posterior has the form $p(\boldsymbol{\theta}|\boldsymbol{D}) = \mathcal{N}_{d}(\boldsymbol{\theta}|\boldsymbol{\mu},\boldsymbol{\Lambda}^{-1})$ where $\boldsymbol{\Lambda}$ denotes the posterior precision matrix (inverse covariance matrix), $\boldsymbol{\mu}$ the posterior mean vector, which are computed as follows:
\begin{align*}
\boldsymbol{\Lambda} &= a\boldsymbol{X}^{T}\boldsymbol{X} + b\boldsymbol{I} \numberthis \\
\boldsymbol{\mu} &= a\boldsymbol{\Lambda}^{-1}\boldsymbol{X}^{T}\boldsymbol{y} \numberthis
\end{align*} 
Thus, for known $a$ and $b$, the posterior of $\boldsymbol{\theta}$ can be easily computed. Even though in real-world applications $a$ is usually not known and a hierarchical model is used instead, the current example is good for testing the utility of our method.

@@

The results on the toy Bayesian regression are depicted in \autoref{fig:Fig.3}. The approximate posterior means show negligible deviations from the analytic posterior means as quantified by very small NRMSEs (as small as $0.002$) and very high $R^{2}$ (as high as $1.0$) over all $n$ sampled during training. This suggests near-perfect estimation of the true posterior means. Further, the estimates become increasingly more accurate, as the number of data points $n$ increases. An inspection of the posterior variances over the different $n$ reveals that the estimated variance follows closely the decrease in analytic variance with increasing $n$. However, the analytic variance is slightly underestimated at smaller $n$ and slightly overestimated at larger $n$. This pattern is also revealed by the KL divergence plot (see \textbf{SI}). This result might be attributable to an underexpressive summary network; another possibility is that the networks need to be trained longer with smaller learning rate decay. 
\begin{figure}[H]
\centering
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Regression_joint.png}
    \caption{Analytic vs. estimated means joint plot ($n=500$)}
    \label{fig:Fig.3a}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
	\begin{subfigure}{.95\textwidth}
 		\includegraphics[width=\textwidth]{Regression_density.png}
    	\caption{Full analytic vs. estimated posteriors ($n=500$)}
    \label{fig:Fig.3b}
	\end{subfigure}
	\begin{subfigure}{.95\textwidth}
 		\includegraphics[width=\textwidth]{Regression_scatter.png}
    	\caption{Analytic means recovery ($n=500$)}
    \label{fig:Fig.3c}
	\end{subfigure}
\end{subfigure}
\begin{subfigure}{.47\textwidth}
 		\includegraphics[width=\textwidth]{Regression_metrics.png}
    	\caption{Performance over all $n$}
    \label{fig:Fig.3d}
\end{subfigure}
\begin{subfigure}{.52\textwidth}
 		\includegraphics[width=\textwidth]{Regression_variance.png}
    	\caption{Analytic vs. estimated variance over all $n$}
    \label{fig:Fig.3e}
\end{subfigure}
\caption[short]{Results on the Bayesian toy regression example. \textbf{(a)} Pair-plots of the analytic vs. estimated posterior means on the whole validation sample. We observe a near-perfect overlap between analytic and estimated posterior means and no spurious covariances; \textbf{(b)} Some example draws from the estimated posteriors. Visual inspections reveals a close match between analytic and estimated posteriors; \textbf{(c)} Posterior mean recovery is almost perfect at $n=500$; \textbf{(d)} NRMSE and $R^{2}$ over all $n$. Performance improves with increasing number of data points; \textbf{(e)} Analytic vs. estimated variance over all $n$. The analytic variance is slightly underestimated at lower $n$ and slightly overestimated at higher $n$.} \label{fig:Fig.3}
\end{figure}

@@

The results on the Ricker model are depicted in \autoref{fig:Fig.4}. As evident from the graphs, parameter recovery becomes better when more time points with data are available (\autoref{fig:Fig.4c}). At $T=500$, the NRMSEs range between $0.015$ and $0.063$, and the $R^{2}$ metrics between $0.997$ and $0.952$, indicating very good recovery of the posterior means (\autoref{fig:Fig.4a}). The parameter $\sigma$ seems to be the hardest to recover. Inspecting the full posteriors, we further see that the posterior distribution of the dummy noise variable $u$ closely resembles the prior, as expected due to the complete lack of mutual information between the data $x$ and $u$ (\autoref{fig:Fig.4b}). Finally, the plots of the rank statistic computed for SCB suggest no systematic distortions of the posterior across all parameters (\autoref{fig:Fig.4d}). Interpreting deviations from uniformity according to \cite{talts2018validating}, the approximate posteriors of $r$ and $\rho$ slightly underestimate the true posterior means, whereas the approximate posterior of $\sigma$ tends to overestimate the true posterior variance. These deviations appear to be due to the fact that recovery worsens at extreme values of the parameters. This is unsurprising, as the data generated with these parameters is highly implausible, which in some cases might even render a model unidentifiable.

@@

As a last example for our method, we use simulations from the \textit{Splat} model and attempt to recover the data-generating parameters. This is a challenging task for any Bayesian method, as the data-generating mechanism is a complicated algorithm, and no likelihood function for joint estimation of the parameters is available. We simulate $200\ 000$  \textit{Gene x Cell} count matrices with different parameter settings. Each count matrix contains the information about $40$ genes sequenced in $100$ different cells, that is, each $\boldsymbol{X} \in \mathbb{R}^{40\times100}$. The parameters considered for estimation, along with a short description, are listed in \autoref{table:Table 1}. The parameter estimation task is thus to recover $\boldsymbol{\theta} = (\alpha,\beta,\mu^{L},\sigma^{L},\pi^{O},\mu^{O},\sigma^{O},\phi)$ from an observed count matrix $\boldsymbol{X}$. We train the networks for 30 epochs through the entire dataset and evaluate the performance on a separate validation set of $300$ count matrices. To stabilize training and avoid exploding gradients due to large counts in the input matrices, we apply a log transformation to each $\boldsymbol{X}$. Simulating the data took approx. $8$ hours. Training the networks took approx. one day with the reference table approach. Inference on $300$ datasets with $2000$ posterior samples per parameter took approx. $1.7$ seconds. 
\begin{figure}[H]
\centering
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_plot.png}
    \caption{The \textit{Splat} simulation diagram. Adapted after \cite{zappia2017splatter}.}
    \label{fig:Fig.7a}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_scatter.png}
    \caption{Parameter recovery}
    \label{fig:Fig.7b}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_sbc.png}
    \caption{Simulation-based calibration (SBC)}
    \label{fig:Fig.7c}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_density.png}
    \caption{Example posterior samples}
    \label{fig:Fig.7d}
\end{subfigure}

\caption[short]{Results on the \textit{Splat} scRNA simulation model. \textbf{(a)} The \textit{Splat} simulation diagram; \textbf{(b)} Parameter recovery; \textbf{(c)} Simulation-based calibraton. The SCB plots indicate a slight underestimation of the true posterior variance.  \textbf{(d)} Example full parameter posteriors from three datasets.} \label{fig:Fig.7}
\end{figure}
The results on the scRNA example are depicted in \autoref{fig:Fig.7}. First, recovery of most parameters is very good, with maximum $R^{2}$ of $0.98$ and minimum NRMSE of $0.039$ for the library location parameter $\mu^{L}$. Interestingly, the rate parameter $\beta$ appears to be nearly unrecoverable, as indexed by the scatterplots in \autoref{fig:Fig.7b}. Furthermore, the estimated posterior of $\beta$ is often bimodal, covering the whole prior range (\autoref{fig:Fig.7c}). This might indicate that the parameter is not identifiable, or that some multivariate trade-offs exist in the parameter space of the \textit{Splat} model. The latter is also confirmed by observing a few marked outliers in the scatterplots of $\alpha$, $\sigma^{O}$, and $\pi^{O}$. Finally, the SCB histograms indicate that our method tends to slightly underestimate the true posterior variance of the $\alpha$, $\beta$, $\mu^{O}$ $\pi^{O}$, and $\sigma^{O}$ parameters. This issue might be due to the fact that the networks were trained with the reference-table approach and might disappear when employing the active learning approach.

@@

In the current work, we proposed and explored a novel end-to-end likelihood-free method which uses invertible neural networks to perform approximate Bayesian inference on any mathematical process model. We demonstrated the utility of the method by applying it to models from various scientific domains dealing with several different data formats and data-generating mechanisms. Further, we explored two possible training approaches suitable for different simulation scenarios, namely, an active learning approach, and a reference-table approach. Both training approaches lead to excellent recovery of the true parameters throughout the examples considered in the current work.

@@

Inspired by previous machine learning approaches to likelihood-free inference \cite{radev2019towards, hwang2018conditional, mestdagh2018prepaid, raynal2018abc, jiang2017learning, lueckmann2017flexible, papamakarios2016fast}, our method shares many of the advantages of these methods and further overcomes some of their important limitations.

First, the introduction of separate summary and invertible networks renders the method independent of the shape or the size of the observed data. The summary network learns a fixed-size vector representation of the data in an automatic, data-driven manner. Since the summary network is optimized jointly with the invertible network, the learned data representation is encouraged to be maximally informative for parameter inference. This is particularly useful in settings where appropriate summary statistics are not known and, as a consequence, relevant information is lost through the choice of suboptimal summary function. However, if highly informative or even \textit{sufficient} statistics are available in a given domain, one might dispose with the summary network altogether and feed these statistics directly to the invertible network.

Second, we showed that the ML loss directly maximizes the posterior over model parameters of interest. This is in contrast to ELBO-based methods which optimize a lower-bound on the posterior \cite{papamakarios2016fast, kingma2014auto}. Therefore, posterior inference is exact when the ML is globally minimized \cite{kingma2018glow, dinh2016density}. This theoretical claim is further confirmed by our results. Since researchers are often interested in some summary of the posterior (e.g., posterior means or variances), we also showed that our method exhibits excellent recovery of the posterior means throughout all examples. Further, parameter recovery becomes better with increasing number of observations, whereas the posterior variance of the parameters decreases. These are important and highly desirable properties of any Bayesian parameter estimation method, as it mirrors the increase in information following increasing number of observations.

Third, the largest computational cost of our method is paid during training. Once trained, the networks can be used and reused to perform inference on large numbers of datasets within seconds and across a given research domain. Indeed, there are many instances of research domains where a single model is extensively explored and independently fitted by multiple researches to test scientific hypotheses \cite{voss2019sequential, zappia2017splatter, ratcliff2008diffusion, de2002fitting}. These research domains are expected to benefit the most from learning the \textit{model universe} once and then inverting the model multiple times for fast inference on multiple datasets. In this regard, our method is similar to the recently introduced prepaid method \cite{mestdagh2018prepaid} which uses a database of pre-computed summary statistics and nearest-neighbors for inference. Note, however, that our method does not need to store training data, since the \textit{knowledge} about the relationship between data and parameters is compressed into the networks' weights. This not only makes the global sharing of pre-trained parameter estimation networks across researchers extremely easy, but also makes their local storage very efficient. Moreover, all computations involved in our method benefit from a high degree of parallelism and can thus utilize the advantages of modern GPUs.

These advantages notwithstanding, some limitations of our method deserve mention. Even though high-level deep learning libraries, such as \textit{TensorFlow} or \textit{Torch}, allow for rapid and relatively straightforward development of various neural network architectures, the implementational burden associated with the current method is still reasonably high. In order to ease the understanding and independent application of the method, we provide fully functioning code to reproduce and study all of the examples tackled in this paper (\href{https://github.com/stefanradev93/cINN}{https://github.com/stefanradev93/cINN}). In addition, we also provide implementation of all tools for performance validation used throughout the paper. Moreover, we are currently developing a general user-friendly software, which should abstract away most of the methodological complexities from the user. Another potential shortcoming of the method is the seemingly overwhelming number of hyperparameters that might require fine-tuning by the user for optimal performance on a given task. However, we observe that many of the default hyperparameter values are sufficient to achieve excellent performance, and starting with a relatively large default network of $10$ ACBs does not appear to hurt performance or destabilize training, even if the mathematical model to be learned is relatively simple. We expect that a single architecture should be able to perform well on almost all models from a given domain (i.e., a single architecture for decision-making models). Future research should investigate the question of generality by applying the method to challenging parameter estimation tasks across different research domains.

>>>>

To assess the performance of our method on the following application examples, we consider a number of different metrics. To assess the precision of point estimates (posterior means), we compute the normalized root mean squared error (NRMSE) and the coefficient of determination ($R^{2}$) between estimated and true parameter values. To assess the recovery of the full posterior, we compute the Kullback-Leibler (KL) divergence \cite{hershey2007approximating} between the true and the approximate distributions for the toy example, and use simulation-based calibration (SBC, \cite{talts2018validating}) for the other examples where the analytic posterior is not available in closed-form. Details for computing the NRMSE, $R^2$, and SCB can be found in the \textbf{SI}. Code for reproducing the results on all following examples, as well as for all network implementations, is freely available at: \href{https://github.com/stefanradev93/cINN}{https://github.com/stefanradev93/cINN}.

@@

As a proof-of-concept, we demonstrate the utility of our method in recovering the true analytic posteriors of the regression coefficients of a conjugate Bayesian regression model. To set the stage, assume we have observed a dataset $\boldsymbol{D} = \{(\boldsymbol{x}^{(i)},y^{(i)})\}_{i=1}^{n}$ with $\boldsymbol{x} \in \mathbb{R}^{d}$ and $y \in \mathbb{R}$. We stack all $\boldsymbol{x}^{(i)}$ row-wise in a design matrix $\boldsymbol{X}$ and model the vector of outcomes $\boldsymbol{y}$ as being conditionally Gaussian given $\boldsymbol{X}$, that is, $\boldsymbol{y} \sim \mathcal{N}_{n}(\boldsymbol{X}\boldsymbol{\theta},a^{-1}\boldsymbol{I})$ where $\boldsymbol{\theta} \in \mathbb{R}^{d}$ is the vector of regression coefficients and $a$ is the precision (inverse noise variance, $a \equiv 1/\sigma_{y}^{2}$). We place a $d$-dimensional diagonal Gaussian prior on the regression coefficients centered at $0$: $\boldsymbol{\theta} \sim \mathcal{N}_{d}(\boldsymbol{0},b^{-1}\boldsymbol{I})$ where $b$ is the precision of the prior ($b \equiv 1/\sigma_{\boldsymbol{\theta}}^{2}$). Thus, the likelihood $p(\boldsymbol{D}|\boldsymbol{\theta})$ admits the following proportionality:
\begin{align*}
p(\boldsymbol{D}|\boldsymbol{\theta}) \propto \exp\left(-\frac{a}{2}\left(\boldsymbol{y}-\boldsymbol{X}\boldsymbol{\theta}\right)^{T}\left(\boldsymbol{y} - \boldsymbol{X}\boldsymbol{\theta}\right)\right) \numberthis 
\end{align*}
Since the prior of $\boldsymbol{\theta}$ is conjugate to the likelihood (both are Gaussian distributions), the posterior of $\boldsymbol{\theta}$ is also Gaussian, given by:
\begin{align*}
p(\boldsymbol{\theta}|\boldsymbol{D}) \propto \exp\left(-\frac{a}{2}\left(\boldsymbol{y}-\boldsymbol{X}\boldsymbol{\theta}\right)^{T}\left(\boldsymbol{y} - \boldsymbol{X}\boldsymbol{\theta}\right)-\frac{b}{2}\boldsymbol{\theta}^{T}\boldsymbol{\theta}\right) \numberthis 
\end{align*}
Therefore, the posterior has the form $p(\boldsymbol{\theta}|\boldsymbol{D}) = \mathcal{N}_{d}(\boldsymbol{\theta}|\boldsymbol{\mu},\boldsymbol{\Lambda}^{-1})$ where $\boldsymbol{\Lambda}$ denotes the posterior precision matrix (inverse covariance matrix), and $\boldsymbol{\mu}$ denotes the posterior mean vector. The posterior moments are computed as follows:
\begin{align*}
\boldsymbol{\Lambda} &= a\boldsymbol{X}^{T}\boldsymbol{X} + b\boldsymbol{I} \numberthis \\
\boldsymbol{\mu} &= a\boldsymbol{\Lambda}^{-1}\boldsymbol{X}^{T}\boldsymbol{y} \numberthis
\end{align*} 
Thus, for known $a$ and $b$, the posterior of $\boldsymbol{\theta}$ can be easily computed. Even though in real-world applications $a$ is usually not known and a hierarchical model is used instead, the current example is well suited for testing the utility of our method.

@@

The results on the toy Bayesian regression are depicted in \autoref{fig:Fig.3}. The approximate posterior means show negligible deviations from the analytic posterior means as quantified by very small NRMSEs (as small as $0.002$) and very high $R^{2}$ (as high as $1.0$) over all $n$ sampled during training. This suggests near-perfect estimation of the true posterior means. Further, the estimates become increasingly more accurate, as the number of data points $n$ increases. An inspection of the posterior variances over the different $n$ reveals that the estimated variance follows closely the decrease in analytic variance with increasing $n$. However, the analytic variance is slightly underestimated at smaller $n$ and slightly overestimated at larger $n$. This pattern is also revealed by the KL divergence plot (see \textbf{SI}). This result might be attributable to an underexpressive summary network; another possibility is that the networks need to be trained longer with smaller learning rate decay. 
\begin{figure}[H]
\centering
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Regression_joint.png}
    \caption{Analytic vs. estimated means joint plot ($n=500$)}
    \label{fig:Fig.3a}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
	\begin{subfigure}{.95\textwidth}
 		\includegraphics[width=\textwidth]{Regression_density.png}
    	\caption{Full analytic vs. estimated posteriors ($n=500$)}
    \label{fig:Fig.3b}
	\end{subfigure}
	\begin{subfigure}{.95\textwidth}
 		\includegraphics[width=\textwidth]{Regression_scatter.png}
    	\caption{Analytic means recovery ($n=500$)}
    \label{fig:Fig.3c}
	\end{subfigure}
\end{subfigure}
\begin{subfigure}{.47\textwidth}
 		\includegraphics[width=\textwidth]{Regression_metrics.png}
    	\caption{Performance over all $n$}
    \label{fig:Fig.3d}
\end{subfigure}
\begin{subfigure}{.52\textwidth}
 		\includegraphics[width=\textwidth]{Regression_variance.png}
    	\caption{Analytic vs. estimated variance over all $n$}
    \label{fig:Fig.3e}
\end{subfigure}
\caption[short]{Results on the Bayesian toy regression example. \textbf{(a)} Pair-plots of the analytic vs. estimated posterior means on the whole validation sample. We observe a near-perfect overlap between analytic and estimated posterior means and no spurious covariances; \textbf{(b)} Some example draws from the estimated posteriors. Visual inspections reveals a close match between analytic and estimated posteriors; \textbf{(c)} Posterior mean recovery is almost perfect at $n=500$; \textbf{(d)} NRMSE and $R^{2}$ over all $n$. Performance improves with increasing number of observed data points; \textbf{(e)} Analytic vs. estimated variance over all $n$. The analytic variance is slightly underestimated at lower $n$ and slightly overestimated at higher $n$.} \label{fig:Fig.3}
\end{figure}

@@

The results on the Ricker model are depicted in \autoref{fig:Fig.4}. As evident from the graphs, parameter recovery becomes better when more time points with data are available (\autoref{fig:Fig.4c}). At $T=500$, the NRMSEs range between $0.015$ and $0.063$, and the $R^{2}$ metrics between $0.997$ and $0.952$, indicating very good recovery of the posterior means (\autoref{fig:Fig.4a}). The parameter $\sigma$ seems to be the hardest to recover. Inspecting the full posteriors, we further see that the posterior distribution of the dummy noise variable $u$ closely resembles the prior. This is expected due to the complete lack of mutual information between the observed data $x$ and $u$ (\autoref{fig:Fig.4b}). Finally, the plots of the rank statistic computed for SCB suggest no systematic distortions of the posterior across all parameters (\autoref{fig:Fig.4d}). Interpreting deviations from uniformity according to \cite{talts2018validating}, the approximate posteriors of $r$ and $\rho$ slightly underestimate the true posterior means, whereas the approximate posterior of $\sigma$ tends to overestimate the true posterior variance. These deviations appear to be due to the fact that recovery worsens at extreme values of the parameters. This is unsurprising, as the data generated with these parameters is highly implausible, which in some cases might even render a model unidentifiable.

@@

As a last example for our method, we use simulations from the \textit{Splat} model to recover the data-generating parameters. This is a challenging task for any Bayesian method, as the data-generating mechanism is a complicated algorithm with no available likelihood for joint estimation of the parameters. Moreover, the observed data exhibits multiple local and non-local dependencies. We simulate $200\ 000$  \textit{Gene x Cell} count matrices with different parameter settings. Each entry in a count matrix contains the number of times a gene has been sequenced in a different cell. We simulate counts for $40$ genes and $100$ cells, that is, each $\boldsymbol{X} \in \mathbb{R}^{40\times100}$. The parameters considered for estimation, along with a short description, are listed in \autoref{table:Table 1}.

The parameter estimation task is thus to recover $\boldsymbol{\theta} = (\alpha,\beta,\mu^{L},\sigma^{L},\pi^{O},\mu^{O},\sigma^{O},\phi)$ from an observed count matrix $\boldsymbol{X}$. We train the networks for 30 epochs through the entire dataset and evaluate the performance on a separate validation set of $300$ count matrices. To stabilize training and avoid exploding gradients due to large counts in the input matrices, we apply a log transformation to each $\boldsymbol{X}$. Simulating the data took approx. $8$ hours. Training the networks took approx. one day with the reference table approach. Inference on $300$ datasets with $2000$ posterior samples per parameter took approx. $1.7$ seconds. 
\begin{figure}[H]
\centering
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_plot.png}
    \caption{The \textit{Splat} simulation diagram. Adapted after \cite{zappia2017splatter}.}
    \label{fig:Fig.7a}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_scatter.png}
    \caption{Parameter recovery}
    \label{fig:Fig.7b}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_sbc.png}
    \caption{Simulation-based calibration (SBC)}
    \label{fig:Fig.7c}
\end{subfigure}
\begin{subfigure}{.49\textwidth}
    \includegraphics[width=\textwidth]{Splatter_density.png}
    \caption{Example posterior samples}
    \label{fig:Fig.7d}
\end{subfigure}

\caption[short]{Results on the \textit{Splat} scRNA simulation model. \textbf{(a)} The \textit{Splat} simulation diagram; \textbf{(b)} Parameter recovery; \textbf{(c)} Simulation-based calibraton. The SCB plots indicate a slight underestimation of the true posterior variance.  \textbf{(d)} Example full parameter posteriors from three datasets.} \label{fig:Fig.7}
\end{figure}
The results on the scRNA example are depicted in \autoref{fig:Fig.7}. First, recovery of most parameters is very good, with maximum $R^{2}$ of $0.98$ and minimum NRMSE of $0.039$ for the library location parameter $\mu^{L}$. Interestingly, the rate parameter $\beta$ appears to be nearly unrecoverable, as indexed by the scatterplots in \autoref{fig:Fig.7b}. Furthermore, the estimated posterior of $\beta$ is often bimodal, covering the whole prior range (\autoref{fig:Fig.7c}). This might indicate that the parameter is not identifiable, or that some multivariate trade-offs exist in the parameter space of the \textit{Splat} model. The latter is also confirmed by observing a few marked outliers in the scatterplots of $\alpha$, $\sigma^{O}$, and $\pi^{O}$. Finally, the SCB histograms indicate that our method tends to slightly underestimate the true posterior variance of the $\alpha$, $\beta$, $\mu^{O}$ $\pi^{O}$, and $\sigma^{O}$ parameters. This issue might be due to the fact that the networks were trained with the reference-table approach and might disappear if the active learning approach is employed.

@@

In the current work, we proposed and explored a novel method which uses invertible neural networks to perform approximate Bayesian inference on any mathematical process model. The method requires only simulations from a model to learn a probabilistic mapping between data and parameters. We demonstrated the utility of the method by applying it to models and data from various scientific domains. Further, we explored two possible training approaches suitable for different simulation scenarios, namely, an active learning approach, and a reference-table approach. Both training approaches lead to excellent parameter estimation throughout the examples considered in the current work.

@@

Inspired by previous machine learning approaches to likelihood-free inference \cite{radev2019towards, hwang2018conditional, mestdagh2018prepaid, raynal2018abc, jiang2017learning, lueckmann2017flexible, papamakarios2016fast}, our method shares many of the advantages of these methods and, further, overcomes some of their important limitations.

First, the introduction of separate summary and invertible networks renders the method independent of the shape or the size of the observed data. The summary network learns a fixed-size vector representation of the data in an automatic, data-driven manner. Since the summary network is optimized jointly with the invertible network, the learned data representation is encouraged to be maximally informative for parameter estimation. This is particularly useful in settings where appropriate summary statistics are not known and, as a consequence, relevant information is lost through the choice of suboptimal summary functions. However, if highly informative or even \textit{sufficient} statistics are available in a given domain, one might dispose with the summary network altogether and feed these statistics directly to the invertible network.

Second, we showed that the ML loss directly maximizes the posterior over the target model parameters without any assumptions on the shape of the posterior. This is in contrast to ELBO-based methods which optimize a lower-bound on the posterior \cite{papamakarios2016fast, kingma2014auto}, and usually assume Gaussian posteriors. Therefore, posterior inference is exact when the ML is globally minimized \cite{kingma2018glow, dinh2016density}. Our results support this theoretical claim. Further, since researchers are often interested in some summary of the posterior (e.g., posterior means or variances), we also showed that our method exhibits excellent recovery of the posterior means throughout all examples. Further, parameter recovery becomes better with increasing number of observations, whereas the posterior variance of the parameters decreases. These are important and highly desirable properties of any Bayesian parameter estimation method, as it mirrors the increase in information following increasing number of observations.

Third, the largest computational cost of our method is paid during training. Once trained, the networks can be used and reused to perform inference on large numbers of datasets within seconds and across a given research domain. Indeed, there are many instances of research domains where a single model is extensively explored and independently fitted by multiple researches to test scientific hypotheses \cite{voss2019sequential, zappia2017splatter, ratcliff2008diffusion, de2002fitting}. These research domains are expected to benefit the most from learning the \textit{model universe} once and then inverting the model multiple times for fast inference on multiple datasets. In this regard, our method is similar to the recently introduced prepaid method \cite{mestdagh2018prepaid} which uses a database of pre-computed summary statistics and nearest-neighbors for inference. Note, however, that our method does not need to store training data, since the \textit{knowledge} about the relationship between data and parameters is compressed into the networks' weights. This not only makes the global sharing of pre-trained parameter estimation networks across researchers extremely easy, but also makes their local storage very efficient. Finally, all computations involved in our method benefit from a high degree of parallelism and can thus utilize the advantages of modern parallel computing paradigms.

These advantages notwithstanding, some limitations of our method deserve mention. Although high-level deep learning libraries, such as \textit{TensorFlow} or \textit{Torch}, allow for rapid and relatively straightforward development of various neural network architectures, the implementational burden associated with the current method is still reasonably high. In order to ease the understanding and independent application of the method, we provide fully functioning code to reproduce and study all of the examples tackled in this paper (\href{https://github.com/stefanradev93/cINN}{https://github.com/stefanradev93/cINN}). In addition, we provide implementation of all metrics and visualization tools for performance validation used throughout the paper. Moreover, we are currently developing a general user-friendly software, which should abstract away most of the methodological complexities from the user. Another potential shortcoming of the method is the seemingly overwhelming number of hyperparameters that might require fine-tuning by the user for optimal performance on a given task. However, we observe that many of the default hyperparameter values are sufficient to achieve excellent performance, and starting with a relatively large default network of $10$ ACBs does not appear to hurt performance or destabilize training, even if the mathematical model to be learned is relatively simple. Based on our results, we expect that a single architecture should be able to perform well on almost all models from a given domain. Future research should investigate the question of generality by applying the method to challenging parameter estimation tasks across different research domains.

We hope that the new method will enable researchers from a variety of fields to accelerate model-based inference and will further prove its utility beyond the examples considered in this paper.
